---
title: "Assignment1"
output: html_document
Name: Eliya Tiram
---

```{r}
library(GGally)
install.packages("data.table")
library(data.table)
library(car)
library(rpart)
library(chemometrics)
```

Preparing the data in the environment

```{r}
# Clear plots
if(!is.null(dev.list())) dev.off()

# Clean workspace
rm(list=ls())

#load data
df <- read.csv("insurance.csv")
```

Data cleaning

```{r}
is.null(df) #no nulls in the data
replace(ddf,which(ddf %like% " "),'')
replace(df,which(df %like% " "), '') #close all blank spaces
which(df=="") #no blanks found in the data
#check for distinct values and whether there are differences in them

unique(df$sex) #expecting 2 values
unique(df$smoker) #expecting 2 values
unique(df$region) #expecting 4 values
#we can see that data is consistent for categorial variables

df$f.sex <- factor(df$sex,labels = c("female","male"));
df$f.smoker <- factor(df$smoker,labels = c("no","yes"))
df$f.region <- factor(df$region,labels = c("northeast","northwest","southeast","southwest"))

summary(df) #from the summary we can see the factor values, it seems that sex and region are distributed equally and not much smokers compare to the non smokers.
dim(df)
unique(df)
#There is only one observation which repeat twice, it makes sense that a person with the same properties will have the same charge and since it's only one I decide to leave it there.

#outliers
par(mfrow=c(1,1))
Boxplot(df$charges)
Boxplot(df$bmi)
#we can see extreme outliers for both charges and bmi, since it's just serval observation it might be the case that for a certain bmi, age or smokers the charge value is raising by a lot compare to the rest. from looking at the high value of column charges it can be seen that all are smokers and mid-high bmi, also some of the ages I see are relatively high.
#infering from graphs
res.out<-Moutlier(df[,c(7,3)],quantile=0.975)
str(res.out)
plot(df$charges,df$bmi)
```

Explanatory data analysis

```{r}
summary(df)
#numeric variables
summary(df[,c(3,7)]) 
plot(df[,c(3,7)])
ggpairs(df[,c(3,7)])

#categorical variables
summary(df[,c(1,4,8:10)])
#from the summary we can see the factor values, it seems that sex and region are distributed equally and not much smokers compare to the non smokers. age and number of children looks about right and there is values in a range that makes sense.


```

-    Determine if the response variable (charges) has an acceptably normal distribution.

```{r}
# Density plot to check the distribution
ggpubr::ggdensity(df$charges,  fill = "lightgray", add = "mean",  xlab = "charges variable density")


# Shapiro Test to asses that data on response variable is normaly distribution
# H0 = Data is normally distributed
# H1 = Data is not normally distributed
# alfa = 0.05
shapiro.test(df$charges)


```

As we can see, the density plot shows that data is not normally distributed. To asses that, we can use one of many statistical tests that check normality on data. In this case, we use Shapiro test.

The result of the Shapiro test shows that data in variable **charges** is not normally distributed since *p-value* is less than the significance level (0.05) so we reject the null hypothesis (data is normally distributed) and we conclude that data is not normally distributed (alternative hypothesis)
