---
title: "Assignment1"
output: html_document
Name: Eliya Tiram
---

# DATA PREPARATION

• ~~Removing duplicate or irrelevant observations [**Eliya**]~~

• ~~Fix structural errors (usually coding errors, trailing blanks in labels, lower/upper case consistency, etc.). [**Eliya**]~~

~~• Check data types. Dates should be coded as such and factors should have level names (if possible, levels have to be set and clarify the variable they belong to). This point is sometimes included under data transformation process. New derived variables are to be produced sometimes scaling and/or normalization (range/shape changes to numeric variables) or category regrouping for factors (nominal/ordinal). [**Eliya**]~~

::: {style="color: red;"}
• Filter unwanted outliers. Univariate and multivariate outliers have to be highlighted.Remove register/erase values and set NA for univariate outiers. [**Eliya**]
:::

::: {style="color: red;"}
• Handle missing data: figure out why the data is missing. Data imputation is to be considered when the aim is modelling (imputation has to be validated). [**Achraf**]
:::

• Data validation is mixed of 'common sense and sector knowledge': Does the data make sense? Does the data follow the appropriate rules for its field? Does it prove or disprove the working theory, or bring any insight to light? Can you find trends in the data to help you form a new theory? If not, is that because of a data quality issue? [**Achraf**]

# TASKS

-   ~~Create factors for qualitative variables. [Eliya]~~

-   ~~Determine if the response variable (charges) has an acceptably normal distribution~~ [Achraf]

-   Address tests to discard serial correlation. [Eliya]

-   Detect univariant and multivariant outliers, errors and missing values (if any) and apply animputation technique if needed. [Achraf]

-   Preliminary exploratory analysis to describe the relationships observed has to be undertaken. [Eliya]

-   If you can improve linear relations or limit the effect of influential data, you must consider the suitable transformations for variables. [Achraf]

-   Apart from the original factor variables, you can consider other categorical variables that can be defined from categorized numeric variables. [Eliya]

-   You must take into account possible interactions between categorical and numerical variables. [Eliya]

-   When building the model, you should study the presence of multicollinearity and try to reduce their impact on the model for easier interpretation. [Achraf]

-   You should build the model using a technique for selecting variables (removing no significant predictors and/or stepwise selection of the best models). [Achraf]

-   The validation of the model has to be done with graphs and / or suitable tests to verify model assumptions. [Achraf]

-   You must include the study of unusual and / or influential data. [Achraf]

-   The resulting model should be interpreted in terms of the relationships of selected predictors and its effect on the response variable. [Eliya]

# ASSIGNMENT

```{r}
library(GGally)
install.packages("data.table")
library(data.table)
library(car)
library(rpart)
library(chemometrics)
```

Preparing the data in the environment

```{r}
# Clear plots
if(!is.null(dev.list())) dev.off()

# Clean workspace
rm(list=ls())

#load data
df <- read.csv("insurance.csv")
```

Data cleaning

```{r}
is.null(df) #no nulls in the data
replace(ddf,which(ddf %like% " "),'')
replace(df,which(df %like% " "), '') #close all blank spaces
which(df=="") #no blanks found in the data
#check for distinct values and whether there are differences in them

unique(df$sex) #expecting 2 values
unique(df$smoker) #expecting 2 values
unique(df$region) #expecting 4 values
#we can see that data is consistent for categorial variables

df$f.sex <- factor(df$sex,labels = c("female","male"));
df$f.smoker <- factor(df$smoker,labels = c("no","yes"))
df$f.region <- factor(df$region,labels = c("northeast","northwest","southeast","southwest"))

summary(df) #from the summary we can see the factor values, it seems that sex and region are distributed equally and not much smokers compare to the non smokers.
dim(df)
unique(df)
#There is only one observation which repeat twice, it makes sense that a person with the same properties will have the same charge and since it's only one I decide to leave it there.

#outliers
par(mfrow=c(1,1))
Boxplot(df$charges)
Boxplot(df$bmi)
#we can see extreme outliers for both charges and bmi, since it's just serval observation it might be the case that for a certain bmi, age or smokers the charge value is raising by a lot compare to the rest. from looking at the high value of column charges it can be seen that all are smokers and mid-high bmi, also some of the ages I see are relatively high.
#infering from graphs
res.out<-Moutlier(df[,c(7,3)],quantile=0.975)


str(res.out)
plot(df$charges,df$bmi)
```

Explanatory data analysis

```{r}
summary(df)
#numeric variables
summary(df[,c(3,7)]) 
plot(df[,c(3,7)])
ggpairs(df[,c(3,7)])

#categorical variables
summary(df[,c(1,4,8:10)])
#from the summary we can see the factor values, it seems that sex and region are distributed equally and not much smokers compare to the non smokers. age and number of children looks about right and there is values in a range that makes sense.


```

-    Determine if the response variable (charges) has an acceptably normal distribution.

```{r}
# Density plot to check the distribution
ggpubr::ggdensity(df$charges,  fill = "lightgray", add = "mean",  xlab = "charges variable density")


# Shapiro Test to asses that data on response variable is normaly distribution
# H0 = Data is normally distributed
# H1 = Data is not normally distributed
# alfa = 0.05
shapiro.test(df$charges)


```

As we can see, the density plot shows that data is not normally distributed. To asses that, we can use one of many statistical tests that check normality on data. In this case, we use Shapiro test.

The result of the Shapiro test shows that data in variable **charges** is not normally distributed since *p-value* is less than the significance level (0.05) so we reject the null hypothesis (data is normally distributed) and we conclude that data is not normally distributed (alternative hypothesis)
